DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2023-10-27 manuscript/manuscript.tex
Rendered pdfs

Chronograms --phylogenies with branch lengths proportional to time-- represent key data on timing of evolutionary events for the study of natural processes in many areas of biological research.
Chronograms also provide valuable information that can be used for education, science communication, and conservation policy decisions.
Yet, achieving a high-quality reconstruction of a chronogram is a difficult and resource-consuming task.
Here we present DateLife, a service implemented as an R package and an R Shiny web application available at www.datelife.org, that provides services for efficient and easy discovery, summary, reuse, and reanalysis of node age data mined from a curated database of expert, peer-reviewed, and openly available chronograms.
The main DateLife workflow starts with one or more scientific taxon names provided by a user. Names are processed and standardized to a unified taxonomy, allowing DateLife to run a name match across its local chronogram database that is curated from Open Tree of Life's phylogenetic repository, and extract all chronograms that contain at least two queried taxon names, along with their metadata.
Finally, node ages from matching chronograms are mapped to corresponding nodes from a chosen tree topology using the congruification algorithm. Congruified node ages are used as secondary calibrations to date the chosen topology, with or without initial branch lengths, using different phylogenetic dating methods such as BLADJ, treePL, PATHd8 and MrBayes.
We performed a cross-validation test to compare node ages resulting from a DateLife analysis (i.e, phylogenetic dating using secondary calibrations) to those from the original chronograms (i.e, obtained with primary calibrations), and found that DateLife's node age estimates are consistent with the age estimates from the original chronograms, with the largest variation in ages occurring around topologically deeper nodes.
Results from any software for scientific analysis can only be as good as the data used as input, we highlight the importance of considering the results of a DateLife analysis in the context of the input chronograms.
We encourage the use of DateLife to help increase awareness of the existing disparities among alternative hypotheses of dates for the same diversification events, and to support exploration of the effect of alternative chronogram hypotheses on downstream analyses, providing a framework for a more informed interpretation of evolutionary results.

@@

Word count: 6707

@@

At the time of writing of this manuscript
(Jun 11, 2023),
DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.

@@

\textbf{\emph{Dating a tree without branch lengths.--}}
When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by applying the Branch Length Adjuster (BLADJ; Webb et al., 2008; Webb \& Donoghue, 2005) algorithm, which requires no initial branch lengths.
The algorithm starts by fixing ages for nodes with calibration data upon the given topology. Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram.
This approach has proven useful for ecological analyses that require a phylogenetic time context (Webb et al., 2008).
When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
The BLADJ algorithm requires a root age estimate to run. If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature. If none is provided, DateLife will choose an age for the root so that it can return a dated topology. It will also provide a conspicuous warning message so that users are aware that the root of the chronogram does not have information available in the chronogram database, along with suggestions on how the user can provide an appropriate age for the root.

In the absence of genetic data, BLADJ is a very agnostic way to assign ages to nodes with no available data, as it does not require any assumptions on the underlying model of branch length distribution. It is however common practice in the literature to use a birth-death model to assign ages to nodes with no genetic data (Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012; Rabosky et al., 2018; Smith \& Brown, 2018). To do so, DateLife implements MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), using nodes with published age data as calibration priors on a fixed topology, a simple birth-death model with parameters that can be determined by the user, and no genetic data.

\textbf{\emph{Dating a tree with branch lengths.--}}
Relative branch lengths can provide key information for phylogenetic dating, specifically for nodes without any calibration data available.
While using initial branch length data is the golden standard for phylogenetic dating analyses, estimating trees with branch lengths proportional to substitution rates per site requires obtaining primary data, assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
DateLife implements a workflow to streamline this process by applying open data from the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for input taxa.
By default, BOLD genetic sequences are aligned with MUSCLE (Edgar, 2004) using functions from the msa R package (Bodenhofer et al., 2015). Alternatively, sequences can be aligned with MAFFT (Katoh, Asimenos, \& Toh, 2009), using functions from the ape R package (Paradis et al., 2004).
The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible (Agnarsson \& Miller, 2008), resulting in older internal nodes as compared to other parsimony algorithms (Forest et al., 2005). The parsimony branch lengths are then optimized using Maximum Likelihood, given the alignment, the topology and a simple Jukes-Cantor model, producing a BOLD tree with branch lengths proportional to expected number of substitutions per site. Both parsimony and ML optimizations are done with functions from the \texttt{phangorn} package (Schliep, 2011).
Due to the computing load it requires, the BOLD workflow is currently only supported through DateLife's R package. It is not yet available through the web application.

>>>>

Chronograms --phylogenies with branch lengths proportional to time-- represent key data on timing of evolutionary events for the study of natural processes in many areas of biological research.
Chronograms also provide valuable information that can be used for education, science communication, and conservation policy decisions.
Yet, achieving a high-quality reconstruction of a chronogram is a difficult and resource-consuming task.
Here we present DateLife, a phylogenetic software implemented as an R package and an R Shiny web application available at www.datelife.org, that provides services for efficient and easy discovery, summary, reuse, and reanalysis of node age data mined from a curated database of expert, peer-reviewed, and openly available chronograms.
The main DateLife workflow starts with one or more scientific taxon names provided by a user. Names are processed and standardized to a unified taxonomy, allowing DateLife to run a name match across its local chronogram database that is curated from Open Tree of Life's phylogenetic repository, and extract all chronograms that contain at least two queried taxon names, along with their metadata.
Finally, node ages from matching chronograms are mapped using the congruification algorithm to corresponding nodes on a tree topology, either extracted from Open Tree of Life's synthetic phylogeny or one provided by the user. Congruified node ages are used as secondary calibrations to date the chosen topology, with or without initial branch lengths, using different phylogenetic dating methods such as BLADJ, treePL, PATHd8 and MrBayes.
We performed a cross-validation test to compare node ages resulting from a DateLife analysis (i.e, phylogenetic dating using secondary calibrations) to those from the original chronograms (i.e, obtained with primary calibrations), and found that DateLife's node age estimates are consistent with the age estimates from the original chronograms, with the largest variation in ages occurring around topologically deeper nodes.
Because the results from any software for scientific analysis can only be as good as the data used as input, we highlight the importance of considering the results of a DateLife analysis in the context of the input chronograms.
DateLife can help to increase awareness of the existing disparities among alternative hypotheses of dates for the same diversification events, and to support exploration of the effect of alternative chronogram hypotheses on downstream analyses, providing a framework for a more informed interpretation of evolutionary results.

@@

Word count: 6797

@@

At the time of writing of this manuscript
(Oct 27, 2023),
DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies. It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database. Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.

@@

\textbf{\emph{Dating a tree without branch lengths.--}}
To date a tree topology when branch lengths are unavailable, DateLife implements the Branch Length Adjuster (BLADJ) algorithm (Webb et al., 2008; Webb \& Donoghue, 2005), which only requires a tree topology with no branch lengths and at least two node ages to use as calibrations, one for the tree root and one for any internal node of the topology.
The BLADJ algorithm fixes ages for nodes with calibration data upon the given tree topology. Then, it assigns ages to nodes with no available age information by distributing time evenly between calibrated nodes, minimizing age variance in the resulting chronogram.
This approach has proven useful for ecological analyses that require a phylogenetic time context (Webb et al., 2008).
When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
The BLADJ algorithm requires a root age to run. Users can provide an appropriate root age estimate of their own or one obtained from the literature. If a root age is not provided and there is no information on the age of the root in the chronogram database, DateLife chooses a random age for the root, so that a dated tree topology can be generated with BLADJ. In this case, DateLife will provide a conspicuous warning message, so that users are aware that the root of the chronogram was chosen at random because there was no information available for it in the chronogram database, along with suggestions on how the user can find and provide an appropriate age for the root of the initial topology.

An alternative to BLADJ to date tree topologies in the absence of initial branch lengths that is common practice in the literature is to use a birth-death model to draw branch lengths (Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012; Rabosky et al., 2018; Smith \& Brown, 2018). In addition to the initial tree topology and nodes with age data, these methods require initial values of speciation and extinction rate parameters provided by the user. DateLife implements this approach with MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), using nodes with published age data as calibration priors on nodes of a tree topology with no branch lengths, a simple birth-death model with speciation and extinction rate parameters that are provided by the user, and no genetic data. However, BLADJ is the default option in DateLife, as it does not require any information on diversification rates for the phylogenetic sample to draw from a branch length distribution.

\textbf{\emph{Dating a tree with branch lengths.--}}
Relative branch lengths can provide key information for phylogenetic dating, specifically for nodes without any calibration data available.
While using initial branch length data is the golden standard for phylogenetic dating analyses, estimating trees with branch lengths proportional to substitution rates per site requires obtaining primary data, assembling and curating a homology (orthology) hypothesis, and choosing and implementing a method for phylogenetic inference.
DateLife implements a workflow to streamline this process by applying open data from the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for input taxa.
By default, BOLD genetic sequences are aligned with MUSCLE (Edgar, 2004) using functions from the msa R package (Bodenhofer et al., 2015). Alternatively, sequences can be aligned with MAFFT (Katoh, Asimenos, \& Toh, 2009), using functions from the ape R package (Paradis et al., 2004).
The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible (Agnarsson \& Miller, 2008), resulting in older internal nodes as compared to other parsimony algorithms (Forest et al., 2005). The parsimony branch lengths are then optimized using Maximum Likelihood, given the alignment, the topology and a simple Jukes-Cantor model, producing a BOLD tree with branch lengths proportional to expected number of substitutions per site. Both parsimony and ML optimizations are done with functions from the \texttt{phangorn} package (Schliep, 2011).
Due to the computing load it requires, the BOLD workflow is currently only supported through DateLife's R package. It is not yet available through the web application.
